---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions,echo=FALSE}
library(knitr)
opts_chunk$set(fig.path = "figure/")
```

## Title

## Synopsis


## Data Processing
The dataset is a CSV file that is compressed to a zip file stored in the working directory.  It contains data submitted by an anonymous invidual recording the number of steps taken using an activity monitoring device like Fitbit.  We'll load the data with read.csv() and use unz() to access the zip file without uncompressing it.  To make grouping and summarizing easier, we'll configure dplyr.
```{r loadAndPreprocess, echo=TRUE, message=FALSE}
mydf <- read.csv(unz("activity.zip","activity.csv"),header=TRUE,sep=",")
library(dplyr)
activity <- tbl_df(mydf)
rm("mydf")
```

## Data Exploration
We'll view the size of the activity data frame and data from the first few rows using dim() and head().  With range() and the subtraction operator, we determine the date range is October 2012 to November 2012, a total of 61 days.  Every 5 minutes, the number of steps were recorded and indexed using 5 minute intervals named 0 to 55, 100 to 155, 200 to 255, etc., and 2300 to 2355.  Essentially the hours are named using the 24-hour clock and each interval is a multiple of five from 5 to 55.  There are a total of 288 five minute intervals a day, which the unique() function displays.
```{r exploreData, echo=TRUE}
dim(activity)
head(activity)
range(as.character(activity$date))
as.Date("2012-12-01")-as.Date("2012-10-01")
unique(activity$interval)
```

## What is mean total number of steps taken per day?
The dplyr group_by and summarize functions makes it easy to compute the total number of steps taken per day, when we combine them with the sum() function and ignore missing values withthe na.rm=TRUE argument.
The results are saved to a new R object totalStepsByDay and names() is helpful in making the variable names more descriptive.
A histogram of the total steps will show the data density; how common or rare values are.  A rug in the histogram helps show all the data points under the histogram and provides some fine detail. 
Finally, we calculate and report the mean and median of the total number of steps taken per day using the R function summary().
We'll add a magenta vertical line to the histogram to denote the mean and a black vertical line to denote the median.  The mean total number of steps taken is 9,354.23 and the median is 10,395.00.
```{r totalStepsByDay, echo=TRUE}
totalStepsByDay <- summarize(group_by(activity, date), sum(steps,na.rm=TRUE))
names(totalStepsByDay) <-c("date","total")
options(digits=10)
summary(totalStepsByDay$total)
```

```{r plot1, echo=TRUE}
hist(totalStepsByDay$total, col="aquamarine", main="Histogram - Total Steps per Day", breaks=5)
rug(totalStepsByDay$total)
abline(v = mean(totalStepsByDay$total), col = "magenta", lwd = 4)
abline(v = median(totalStepsByDay$total), col = "black", lwd = 4)
```

## What is the average daily activity pattern?
Using dplyr summarize and group_by, we compute the means of steps for each 5-minute interval and ignore missing values.
We'll plot the mean steps taken by 5-minute interval across all days.  We'll set the digits option to 10 to increase the number of significant digits, when viewing the summary and subset output.
The R summary function tells us the maximum mean number of steps is 206.1698, which we can plug in to the subset function and find the 5-minute interval with this value.
The interval 835 contains the maximum mean number of steps of 206.1698.
```{r meanStepsByInterval, echo=TRUE}
meansByInterval <- summarize(group_by(activity,interval),mean(steps,na.rm=TRUE))
names(meansByInterval) <-c("interval","means")

options(digits=10)
summary(meansByInterval$means)

subset(meansByInterval,means>206)
```

```{r plot2, echo=TRUE}
with(meansByInterval,plot(interval,means,type="l", main="Average number of steps taken"))
```

## Imputing missing values
Values are only missing in the steps variable and the logical vector, missingSteps helps identify the days when used with the unique() function.  The 8 days with missing values are 2012-10-01 2012-10-08 2012-11-01 2012-11-04 2012-11-09 2012-11-10 2012-11-14 2012-11-30.

The function nrow returns 2,304 or the total number of missing values in the dataset when we subset using the missingSteps logical vector.

Since we already have the means by interval for the entire dataset in the meansByInterval R object, my strategy is to set the missing value for each interval to the mean for each interval for each of the eight days.

First, we'll create a new dataset called completeActivity by subsetting activity and selecting rows with dates NOT EQUAL to 2012-10-01 2012-10-08 2012-11-01 2012-11-04 2012-11-09 2012-11-10 2012-11-14 2012-11-30.  This gives us the rows with recorded values for steps.

Then, we create temporary R objects, (add1, add2, add3, add4, add5, add6, add7, and add8) for each day with missing values and apply our strategy and set the missing values for each interval to the mean for each interval in the R object meansByInterval.

We merge each day with imputed values into the new dataset completeActivity and check for missing values in the new dataset and find none.
We do some housekeeping by clearing the temporary R objects, add1 to add8.

To create our dataset for the plot, we use dplyr group_by and summarize to compute total number of steps taken per day using the completeActivity dataset that contains imputed missing values and save in a new R object, totalStepsByDayCompleteActivity and make the variables more descriptive.

We calculate and report the mean and median total number of steps taken per day using the the dataset with imputed missing values and compare to the summary from the original dataset.

We make a histogram of the total number of steps taken each day using the dataset with imputed missing values and compare to the histogram we did earlier.  It is easier to compare the plots, when one is placed on top of the other.  In the new histogram, the mean and median are the same, so we distinguish the mean by making it a wide magenta dotted line.

```{r imputeMissingValues, echo=TRUE}
sum(is.na(activity$date)) # 0 rows
sum(is.na(activity$interval)) # 0 rows
missingSteps <- is.na(activity$steps)
unique(activity[missingSteps,2]) 
nrow(activity[missingSteps,]) # 2,304 rows

completeActivity <- subset(activity,!(date %in% c("2012-10-01","2012-10-08","2012-11-01","2012-11-04","2012-11-09","2012-11-10","2012-11-14","2012-11-30")))

add1 <- data.frame(steps=meansByInterval$means,date="2012-10-01",interval=meansByInterval$interval)
add2 <- data.frame(steps=meansByInterval$means,date="2012-10-08",interval=meansByInterval$interval)
add3 <- data.frame(steps=meansByInterval$means,date="2012-11-01",interval=meansByInterval$interval)
add4 <- data.frame(steps=meansByInterval$means,date="2012-11-04",interval=meansByInterval$interval)
add5 <- data.frame(steps=meansByInterval$means,date="2012-11-09",interval=meansByInterval$interval)
add6 <- data.frame(steps=meansByInterval$means,date="2012-11-10",interval=meansByInterval$interval)
add7 <- data.frame(steps=meansByInterval$means,date="2012-11-14",interval=meansByInterval$interval)
add8 <- data.frame(steps=meansByInterval$means,date="2012-11-30",interval=meansByInterval$interval)

completeActivity <- rbind(completeActivity,add1)
completeActivity <- rbind(completeActivity,add2)
completeActivity <- rbind(completeActivity,add3)
completeActivity <- rbind(completeActivity,add4)
completeActivity <- rbind(completeActivity,add5)
completeActivity <- rbind(completeActivity,add6)
completeActivity <- rbind(completeActivity,add7)
completeActivity <- rbind(completeActivity,add8)

sum(is.na(completeActivity$steps)) # 0 rows

rm("add1");rm("add2");rm("add3");rm("add4");rm("add5");rm("add6");rm("add7");rm("add8");

totalStepsByDayCompleteActivity <- summarize(group_by(completeActivity, date), sum(steps,na.rm=TRUE))
names(totalStepsByDayCompleteActivity) <-c("date","totals")

options(digits=10)
summary(totalStepsByDayCompleteActivity$totals)

summary(totalStepsByDay$total)
```

```{r plot3, echo=TRUE, fig.height=8}
par(mfrow = c(2,1))
hist(totalStepsByDay$total, col="aquamarine", main="Histogram - Total Steps per Day", breaks=5)
rug(totalStepsByDay$total)
abline(v = mean(totalStepsByDay$total), col = "magenta", lwd = 4)
abline(v = median(totalStepsByDay$total), col = "black", lwd = 4)

hist(totalStepsByDayCompleteActivity$totals, col="rosybrown", main="Histogram -Total Steps per Day using Imputed Missing Values", breaks=5)
rug(totalStepsByDayCompleteActivity$totals)

abline(v = mean(totalStepsByDayCompleteActivity$totals), col = "magenta", lwd = 8,lty=3)
abline(v = median(totalStepsByDayCompleteActivity$totals), col = "black", lwd = 4)
```

## Are there differences in activity patterns between weekdays and weekends?
We need a custom function makeDayTypeFactor that takes a vector of dates and returns a vector of factor variables with levels weekday and weekend, using the weekdays() function.  This can be used to create a new dayType variable in the completeActivity dataset that has imputed missing values.

With the new factor variable, we use dplyr group_by and summarize to compute mean number of steps taken, by interval and day type and save in the new R object meansByIntervalByDayTypeCompleteActivity and make the variable names more descriptive.

Finally, we use lattice to make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekdays or weekends.
```{r patternsByDayType, echo=TRUE}
makeDayTypeFactor <- function(v) {
    results <- vector()
    for (s in v) {
        if (weekdays(strptime(s,"%Y-%m-%d")) %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")) {
            results <- c(results,"weekday")
        } else {
            results <- c(results,"weekend")
        }        
    }
    as.factor(results)
}

completeActivity$dayType <- makeDayTypeFactor(completeActivity$date)

meansByIntervalByDayTypeCompleteActivity <- summarize(group_by(completeActivity, interval, dayType), mean(steps))

names(meansByIntervalByDayTypeCompleteActivity) <-c("interval","dayType", "means")
```

```{r plot4, echo=TRUE}
library(lattice)
xyplot(means ~ interval | dayType, type="l", data=meansByIntervalByDayTypeCompleteActivity, layout = c(1,2))
```

## Results
### What is mean total number of steps taken per day?
The mean total number of steps taken per day is 9,354.23 with a median of 10,395.00.  The 1st Quartile is 6,778, 3rd Quartile is 12,811, and a maximum of 21,194.  The histogram shows total steps concentrated in the range 10,000 to 15,000.  There are more data values in the range 0 to 10,000, than in the range 15,000 to 25,000.

### What is the average daily activity pattern?
The largest mean number of steps taken within a day's 5-minute interval is 206.1698 in the 835 interval or at 8:35 am.  From intervals 0 to 500, the average number of steps is usually 0.  From time intervals 500 to about 900, the average number of steps increase from 0 to about 200.  There seems to be a pattern for the average number of steps for the intervals 1000 to 2000.  The average moves from around 50 to 100 and back to 50, during the ranges, 1000 to 1500, 1500 to 1750, and 1750 to 2000.  From intervals 2000 to 2355, the average moves gradually from 50 to 0.

### Do these values differ from the estimates from the first part of the assignment?
This table compares the summaries of the original dataset and the dataset with imputed missing values.
ORIGINAL
     Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
     0.00  6778.00 10395.00  9354.23 12811.00 21194.00
AFTER IMPUTING MISSING VALUES
     Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
    41.00  9819.00 10766.19 10766.19 12811.00 21194.00

In the original dataset, the mean total number of steps taken per day is 9,354.23 with a median of 10,395.00.  The 1st Quartile is 6,778, 3rd Quartile is 12,811, and a maximum of 21,194.  The histogram shows total steps concentrated in the range 10,000 to 15,000.  There are more data values in the range 0 to 10,000, than in the range 15,000 to 25,000.

After imputing missing values, the mean and median total number of steps taken per day is 10,766.19.  The 1st Quartile is 9,819, 3rd Quartile is 12,811, and a maximum of 21,194.  The histogram shows total steps concentrated in the range 7,500 to 15,000.  There are less data values in the ranges 0 to 5,000, and the range 20,000 to 25,000.

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
After imputing missing values, and repeating the same analysis of the total number of steps per day, the new dataset distribution looks more Gaussian or normal.  When we compare the two histograms, the new plot contains more data about the mean.  As a matter of fact, the mean and median are the same after imputing missing data.

### Are there differences in activity patterns between weekdays and weekends?
Yes, on weekends, the average steps taken per day goes to 50 later in the day than on the weekdays; about 750 compared to about 550.
Also, the pattern of the average in the intervals 1000 to 2000 is different on weekends.  The average moved to a high of about 150 on weekends, compared to 100 on weekdays.  There was also a couple of patterns visible during the weekends where the average moved from 50 to 100 and back to 50; the average also moved from 50 to 150 and back to 50.
During weekends and weekdays, the average steps moved from 50 to 0 in the intervals 2000 to 2300.
